Measuring the double-drift illusion with hand tracking

### 2020-12-12 Manuscript.Rmd
implemented some feedback from Patrick

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived direction deviates strongly from its true direction and is instead some combination of its real external motion and its internal motion [@Tse2006]. The perceived location of such a double-drift stimulus periodically resets toward the veridical position. There may indeed be a simple perdiodicity underlying these resets. However, since both the veridical and perceived location are available to the brain, these resets may also occur when these two positions are too far from each other to come from the same source stimulus. Here we test which of the two explanations is more likely. In a first experiment, we had participants continuously track the position of the perceived position of the stimulus, using a stylus on a drawing tablet. While the orientation of the recorded trajectories reflect stimulus properties, there were no spontaneous resets. In a second experiment, we had participants re-trace the perceived trajectory of a short stimulus presentation, or indicate the initial direction of the stimulus they perceived. We found resets here, and they indicate a set spatial limit to the distance between the veridical and perceived location of the double-drift stimulus.

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence its perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor with internal motion orthogonal to its path is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. And although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018]. These findings suggest that both the true and illusory positions are available in the brain.

```{r fig1, echo=F, fig.height=8, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Top left:_ a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path. Eye-movements immediately following disappearance of the gabor land at the true position, but pointing movements and delayed eye-movements go to the illusory position. _Bottom left:_ at some point, the perceived position will 'reset' towards the true position. For some people this takes the form of a 'jump' back to the true position, for others it means the illusory position doesn't deviate more from the true position, and there are some in between version. _Top right:_ if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate. _Middle right:_ if they occur after some period of time, they should have a constant Y coordinate. _Bottom right:_ we have people track or re-trace their percept and try to isolate reset points from the drawn trajectories to figure out where or when resets occur."}
library(magick)
Fig1 <- image_read_svg('doc/stimulus_reset.svg')
par(mai=c(0,0,0,0))
plot(Fig1)
```

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates after about 100 ms [@Chung2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. The strength of the illusion, or how much the perceived position shifts from the real position over time, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015].

At some point, the illusion resets, the perceived location jumps back to its physical location and then accumulation begins again (Fig. 1, bottom left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets. Alternatively, there is a simple periodicity or stochastics process underlies the changes in perception. Recent work has shown that independent of the stimulus, movement-related attention may lead to resets [@Nakayama2020]. This means that participants' responses to the stimulus may change the illusion. Here we will first test if real-time as well as delayed, manual tracking of the gabor's position, prevents or decreases the illusory shift in position.

@@

# Methods

## Participants

For the first task in experiment 1, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). For the second task in experiment 1, 6 participants were recruited from the lab (4 female; ages 19 – 26, mean: 21.6). For experiment 2, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and had either normal or corrected to normal vision. Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

## Setup

Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet. The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus.

For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assumed the participants to be close to the setup, with their eyes at the height of the monitor. Since there was no requirement to use the forehead rest, the latter is probably only true for the tallest participants, so that reported _dva's_ are slightly underestimating the size of stimuli.

## Stimuli

The stimuli were phase-changing gabor patches with a gaussian mask with an sd of 0.43 _cm_ ( or ~.6 _dva_ when in the centre of the screen and viewed from a typical distance ). The whole stimulus extended 3 times the sd of the Guassian in all four cardinal directions. The stimulus used 1.5 phases of a sine wave, so that each sine phase equals ~0.58 _cm_ or ~1.5 _dva_. By applying both real motion and phases changes to the gabor, it becomes a 'double-drift stimulus' where the perceived path deviates from the real path. The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ). The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move forward and backward on the horizontal display. This real motion was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from either extreme to the other ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of 6.75, 4.5 and 3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ). For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).

## Tasks

In the two tasks in experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus. In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus. More details are given below.

## Analyses

For experiment 1, we first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the first and last half second ( leaving 1 _s_ trajectories for the first experiment, and 3 _s_ trajectories for the second experiment). We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples). This heading measure should depend on the illusion strength.

For comparison, we use the same heading measure in experiment 2. However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect left-right direction changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 _cm_ away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.

Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].

# Experiment 1: Continuous tracking

@@

## Task

In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so. Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 _cps_. Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path.

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass (as tracking was noisy at the inversion points) for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking does not prevent the illusion from ocurring. We will use the averaged angle for each of the internal motion conditions as a measure of the strength of the illusion.

```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2. Heading and Trajectories in Experiment 1a.** **Top:** Average heading across the middle 1 _s_ of each pass. Distance from origin is time, angle is angle. In blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading. **Bottom:** All individual trajectores. _Red_: Highlighted trajectories from conditions with non-zero internal speed, where at least some heading samples fell in the 95% confidence interval for that participants' heading distribution in zero internal speed condition."}
plotBoundedTracking()
```

@@

The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it? We compare illusion strength to that reported previously [@Cavanagh2019] in Figure 3. As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model. That is; tracking the illusion while observing it does not change the strength of the illusion.

```{r fig3, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3. Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ). In red are the average illusions strengths for ±3 _cps_, ±1 _cps and 0 cps. The black circles and line are the data and model from Cavanagh & Tse (2019). The grey lines are the illusion strengths their model predicts for the internal motion speeds used here."}
plotStrengthComparison()
```

## Resets

In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus. Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition. We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset. These trajectories are drawn in red in Figure 2 (for clarity only 1 per condition/participant is highlighted), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.

No unambiguous resets were captured in either experiment. It is possible that resets were perceived but were too quick to respond to in the continuous tracking task. Alternatively, tracking may have stabilized the illusion, preventing resets from occurring. Either way, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.

# Experiment 2: Re-tracing

In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, as used by Nakayama & Holcombe [-@Nakayama2020]. This allows participants more time to carefully reproduce the path. We use the recorded trajectories to assess points where resets occured. The reset points can then be used to determine if resets are time-limited or space-limited.

## Tasks

Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. The re-tracing task was done in half of 8 blocks. In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion. Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times. Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.

@@

@@

```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6. Reset Points and Single Limits** **Left:** True path of the doubel-drift stimulus as gray arrow. The average reset points for each participant and stimulus condition displayed spatially, with the internal motion normalized to go rightward. Y-coordinates have been converted to time ( _s_ ), in order to normalize for the two external speeds. Transparent purple lines indicate the standard deviation of coordinates. Dashed vertical red line indicates the median X-coordinate (1.6 _cm_). Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_). **Top right:** X-coordinates predicted from a single temporal limit. **Bottom right:** Y-coordinates predicted from a single spatial limit."}
plotResetPoints()
```

@@

Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope. If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=258.4) and a spatial limit of ~1.7 _cm_ (AIC=217.7). This is close to the median values of the respective coordinates. The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs). This again indicates that a spatial limit explains more of the data.

@@

The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the MSE and AIC of the model are lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui Liu] obtained the point of subjective equivalence (PSE) of a double-drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 _dva_ length trajectory, traversed in 1, 2 or 3 _s_. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here (see Fig 8).

```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
plotAverageTrajectories()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials are also included in the PSE data. Further, we proportionally scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. Finally, for the 4 _s_ stimuli, we removed the last second. Figure 9 (top row) shows the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) as red dots, with the time-based model's average trajectories in black and the distribution of percepts as a gray histogram. The bottom row of Figure 9 shows the averaged trajectories across participants for each of the internal motions (purple, red, orange), and external motions (left and right), with the distribution of percepts as a gray histogram for the same 3 time points as in the top row.

>>>>

---
title: "Measuring the double-drift illusion with hand tracking"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

If a gabor pattern drifts in one direction while its internal texture drifts in the orthogonal direction, its perceived position deviates further and further away from  its true path. We first evaluated the illusion using manual tracking. Participants followed the gabor with a stylus on a drawing tablet that coincided optically with the horizontal monitor surface. Their hand and the stylus were not visible during the tracking. The magnitude of the tracking illusion corresponded closely to previous perceptual and pointing measures indicating that manual tracking is a valid measure for the illusion. This allowed us to use it to capture the behavior of the illusion as it eventually degrades and breaks down in single trials. Specifically, the deviation of the gabor stops accumulating at some point and either stays at a fixed offset or resets toward the veridical position. In a second experiment, participants re-traced the perceived trajectory after the gabor was removed from the monitor. On alternate trials, they indicate the initial direction of the gabor’s path. Resets were detected and analyzed and they suggest that there is a spatial limit to  the perceived offset beyond which the deviation saturates or resets.

@@

Encoding the position of objects in the world is necessary for the myriad of visually guided motor tasks that we do every day. It is no surprise then that the brain has several mechanisms to gauge and update the position of objects in the world. For example, when a target is moving, its motion can influence its perceived location, making it appear to be slightly ahead of its true location, e.g. the flash lag [@Nijhawan1994]; or the flash grab effect [@Cavanagh2013]. These cases of position extrapolation may be functional, helping the motor system overcome inevitable neural delays in targeting a moving object [@Nijhawan1994; @Duhamel1992]. A stronger and very different motion-induced position shift arises when a moving gabor has internal motion orthogonal to its path and is viewed in the periphery [@Tse2006; @Shapiro2010; @Gurnsey2012; @Kwon2015; @Lisi2015]. This double-drift stimulus (Fig. 1, top left) generates extreme misjudgments of the moving target’s location which may deviate by as much as several degrees of visual angle from its true location. Surprisingly, this extraordinary perceptual illusion does not affect eye movements to the gabor: saccades to these targets are determined by their physical not their perceived location [@Lisi2015]. And although immediate eye-movements go towards the true position, delayed eye movements and pointing go to the perceived position [@Lisi2017; @Massendari2018]. These findings suggest that both the true and illusory positions are available in the brain.

```{r fig1, echo=F, fig.height=3.2, fig.width=8, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 1: Stimuli and Setup.** _Left_: a gabor with internal motion (phase shifts) and external motion, that is viewed peripherally, appears to follow a path deviating from the true path. _Right_: participants track their as the gabor moves back and forth along its path."}
library(magick)
Fig1 <- image_read_svg('doc/methods_fig_1.svg')
par(mai=c(0,0,0,0))
plot(Fig1)
```

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues away from its physical location up to several times the size of the gabor. The strength of the illusion, or how much the perceived position shifts from the real position over time, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015]. In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig. 1) with 4 internal speeds. We analyze the angle of the tracking to determine if manual response is susceptible to the illusion.

At some point, the illusion stops accumulating. The position may just remain at some fixed offset following a path parallel to the true path or the position may reset back toward its physical location and then accumulation begins again (Fig. 4, left) or it saturates. This long duration before resetting or saturation is well beyond the integration period of other visual effects and into the range of short-term memory. Given that both true and illusory positions are available, it could be that instead of some integration period, it is the interplay between the two position signals that leads to spontaneous resets. Alternatively, there is a simple periodicity or a stochastic process that underlies the changes in perception. Recent work has shown that independent of the stimulus, movement-related attention may lead to resets [@Nakayama2020]. This means that participants' responses to the stimulus may change the illusion. Here we will first test if real-time as well as delayed, manual tracking of the gabor's position, prevents or decreases the illusory shift in position.

@@

# Online tracking experiment

@@

## Methods

### Participants

For this experiment, 4 participants were recruited from the lab (3 female; ages 21 – 31, mean: 25.75). All participants reported to be right handed and had either normal or corrected to normal vision. Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

### Setup

Participants used a stylus on a horizontal drawing tablet (Wacom Intuos Pro) to indicate where they perceived the location of a gabor (see Fig 1, bottom right), as well as a keypad for additional responses. Stimuli were displayed on a downward facing LED monitor (30 Hz, 20”, 1680x1050), parallel with the drawing tablet. The stimuli were observed via an upward facing mirror placed exactly between the tablet and monitor, so that the stimuli appear to be in the same horizontal plane as the hand, and the tip of the stylus. Experiments were run in Python 2.7 with PsychoPy [@Peirce2019] and analyses were done in R 3.6.1 [@Rproject].

For reporting degrees of visual angle that stimuli spanned ( _dva_ ), we assumed the participants to be close to the setup, with their eyes at the height of the monitor. While there was no requirement to use the forehead rest, this is a reasonable assumption, since stimuli are only visible when relatively close to the setup. Nevertless, reported _dva's_ are estimates of the size of stimuli.

### Stimuli

The stimuli were phase-changing gabor patches with a gaussian mask with a sigma (sd) of 0.43 _cm_ or ~0.59 _dva_, spatial frequency of 0.58 _c/cm_ or ~0.42 _c/dva_, vertical orientation, and contrast of 100%. By simultaneously moving the gabor envelope and drifting its internal sine wave, it becomes a ‘double-drift stimulus’ where the perceived path deviates from the real path. The gabor changed phase at different rates, called 'internal motion' here, specified in cycles per second ( _cps_ ). The real displacements of the stimulus are along the Y axis of the monitor, keeping the X coordinate constant at the middle of the screen, so that a stimulus without internal motion would appear to move toward and away from the participant on the horizontal display.

The motion of the envelope was restricted to the central 13.5 _cm_ ( ~14.0 _dva_ ) and was set so that the stimulus would move from one end to the other of its path ( a motion we call a 'pass' ) in 2, 3 or 4 _s_, corresponding to speeds of ~6.75, ~4.5 and ~3.75 _cm/s_ ( or roughly 7.0, 4.7 or 3.5 _dva/s_ ). For simplicity, and since the distance moved is always the same, these speeds will be refered to by the duration of a single pass (2, 3 or 4 _s_).

### Procedure

In experiment 1, participants were to do online tracking (using their unseen hand / stylus) while viewing a double-drift stimulus. In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 s, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 s to do so. Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 cps. Since each moving gabor was shown for 12 s, and it moved from one end to another in 2 s, it completed 5 full and 2 half passes of the path.

### Analyses

We first segmented the tracking trajectories according to stimulus passes (in between direction changes of the gabor), and removed the noisy tracking during the first and last half second, leaving 1 _s_ trajectories. We then calculated the instantaneous heading (direction, disregarding velocity or distance) of a smooth spline interpolated version of each trajectory segment (using 101 spline interpolated position samples, to get 100 heading samples). This heading measure should depend on the illusion strength.

@@

In Figure 2 (top), we plot the average instantaneous heading during the middle second of each 2 _s_ pass for each of the 5 internal motion speeds (normalized for the external motion). The average angle of the tracked path is given in degrees deviation from straight forward. It seems that continuous tracking is sensitive to the strength of the illusion, and this is confirmed by a repeated-measures F-test on the average angles, using internal motion as a within-subjects factor ( _F_(4,12)=102.6, _p_<.001, $\eta^2$=0.97 ). With FDR-corrected, paired t-tests, we find a difference in heading between 0 _cps_ internal motion and both 1 _cps_ ( _p_=.022) and -1 _cps_ ( _p_=.025) as well as between -3 and -1 _cps_ and 3 and 1 _cps_ (both _p_ <.001). This makes it clear that continuous tracking does not prevent the illusion from occurring. We will use heading angle as a measure of the strength of the illusion.

We tried detecting resets, but none could be unambiguously identified (see Fig 2). It is possible that resets were perceived but were too quick to respond to in the continuous tracking task. Alternatively, tracking may have stabilized the illusion, preventing resets from occurring.

```{r fig2, echo=FALSE, message=FALSE, fig.width=4.5, fig.height=6, fig.cap="\\label{fig:fig2}**Figure 2. Heading and Trajectories in Experiment 1.** **Top:** Average heading across the middle 1 _s_ of each pass. Distance from origin is time, angle is angle. In darker blue lines the average heading across participants, in light blue a polar heatmap giving an impression of the distribution of heading. **Bottom:** All individual trajectores. _Red_: Highlighted trajectories from conditions with non-zero internal speed, where at least some heading samples fell in the 95% confidence interval for that participants' heading distribution in zero internal speed condition. This would occur during a reset, but only very few trajectories have this feature, and none contain clear resets."}
plotBoundedTracking()
```

@@

The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it? We compare mean illusion strength to that reported previously [@Cavanagh2019] in Figure 3. As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model (in gray). That is; tracking the illusion while observing it does not change the strength of the illusion.

```{r fig3, echo=FALSE, message=FALSE, fig.width=5, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3. Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ). In red are the average illusions strengths for 0 _cps_, ±1 _cps_ and ±3 _cps_, and in light blue the data for individual participants. The gray lines are the predicted illusion strengths based on the model from Cavanagh & Tse (2019) for the internal motion speeds used here. *the second panel only for testing*"}
plotStrengthComparison()
```

# Delayed retracing experiment

In a second experiment, aimed at collecting resets, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, a method used by Nakayama & Holcombe [-@Nakayama2020]. This allows participants more time to carefully reproduce the path. We use the recorded trajectories to assess points where resets occurred. The reset points can then be used to determine if resets are time-limited or space-limited.

```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** _Left_: At some point, the perceived position will stop drifting further and further away from its actual path. For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, and there are some in-between versions. _Top right_: if these resets are due to a spatial limit on the size of the illusory position shift, the points of the resets should have a constant X coordinate. _Bottom right_: if they occur after some period of time, they should have a constant Y coordinate."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

## Methods

### Participants

For this experiment, 9 participants were recruited from the lab (6 female; ages 19 - 27, mean: 22.8). All participants reported to be right handed and had either normal or corrected to normal vision. Procedures were approved by York’s Human Participants Review Committee, and all participants provided prior, written, informed consent.

### Setup and Stimuli

The setup was the same as used in the online tracking experiment, except that we now also used a small keypad, for responses (see Fig 1, right). Stimuli were identical to those used in the online tracking experiment, but now always started at the near end of the workspace and moved away from the participants. The double drift stimuli would use either 3 or 4 seconds to move 13.5 _cm_ and had an internal drift of 2, 3 or 4 _cps_.

## Procedure

In experiment 2, participants first watched a double-drift stimulus and then either reproduced the perceived path of the stimulus by retracing it on the tablet, or indicated the initial movement direction by changing the orientation of a line originating in the same position as the doube-drift stimulus. More details are given below. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. The re-tracing task was done in half of 8 blocks. In the other half, participants viewed the same stimuli, and then oriented a line to indicate the initial perceived direction of the stimulus motion. Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times. Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.

@@

### Analyses

We use the same heading measure here as in experiment 1. However, the individual trajectories reveal more points that are potential spontaneous resets that we try to extract. Using a smooth spline interpolated version of the trajectory we detect left-right direction changes. Since the initial movement is often jittery, we only take direction changes further than ~0.39 cm away from the starting point. The first (if any) of the remaining direction changes, are then taken as reset points. By averaging the x and y coordinates of reset points in all 9 participants and 6 stimulus conditions, we obtain 54 average reset points for further analyses and modelling. The average line orientation is also calculated for each of those 54 combinations of partcipants and stimulus conditions.

@@

```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6. Reset Points and Single Limits** **Left:** True path of the double-drift stimulus as gray arrow. The average reset points for each participant and stimulus condition displayed spatially, with the internal motion normalized to go rightward. Y-coordinates have been converted to time ( _s_ ), in order to normalize for the two external speeds. Transparent purple lines indicate the standard deviation of coordinates. Dashed vertical red line indicates the median X-coordinate (1.6 _cm_). Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_). **Top right:** X-coordinates predicted from a single temporal limit. **Bottom right:** Y-coordinates predicted from a single spatial limit."}
plotResetPoints()
```

@@

Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope. If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=-167.8) and a spatial limit of ~1.7 _cm_ (AIC=-208.6). This is close to the median values of the respective coordinates. The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs). This again indicates that a spatial limit explains more of the data.

@@

The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the AIC of the model (-209.8) is marginally lower than that of the model with the single spatial limit, but it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. An as of yet unpublished study [@Liu2019] obtained data on the PSE of a double-drift stimulus at the end of trajectories of various lengths traveled at different speeds. These PSEs were predicted very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_). Interestingly, this temporal constraint fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then we should see a similar pattern in the averaged trajectories in our re-tracing experiment. However, these averaged trajectories reach a plateau within ~1 _s_ after which resets and illusory position shift seem to balance each other. This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. First, our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias. We can't reolve this contradiction here and leave that for future work.
